The Dynamics of Anthropomorphic Effects in Human-Robot Interaction

### 2014-03-21 paper.tex
refinements Roomba study part

We openly explored long-term interactions between human users and a domestic service robot in a 6-months lasting ethnographic study. We gave a commercially available vacuum cleaning robot (iRobot's Roomba) to nine different types of households in the French speaking part of Switzerland. The fairly diverse sample of households consisted of 2 single homes, 1 couple, and 6 families with children in their home. In total, this summed up to 30 participants: 15 adults (26-71 years old, mean age 43.6 years), and 15 children (6 months to 18 years old). During the 6 months, each household was visited 5 times to conduct interviews and make field observations in the households. Moreover during several weeks, households were asked to fill in a diary to report the usage of the robot and other cleaning activities.

The main goal of this study was to explore how households live with a vacuum cleaning robot over an extended period of time. We wanted to analyze how the usage of and interaction with the robot evolve over time, and to identify adoption phases as well as factors that impact on the long-term acceptance of service robots in homes.
\footnote{This study is published in \cite{fink_living_2013}.} Hereafter, we refer to this study as the "Roomba study".

From the Roomba study, we can report on several findings related to anthropomorphism and the anthropomorphic design of robots. The gathered data for these aspects include qualitative observations, semi-structured interviews, and also quantitative assessments of various aspects of human-robot interaction. The Roomba study is one of very few long-term human-robot interaction studies in the domestic environment.

In the following, we present findings in respect of two aspects:
First, we qualitatively describe what kinds of anthropomorphic interactions between participants and the robot were observed and outline how these interactions evolved over time.
Second, we present people's perception of human-likeness and social capabilities of the robot and how this perception changed over time.

\subsubsection{Anthropomorphic interactions with the robot}

It has been described in several previous studies of the Roomba \citep{forlizzi_service_2006,forlizzi_how_2007,krumm_my_2007,sung_domestic_2010}, that users tend to engage in a variety of non-cleaning activities associated with the Roomba including, but not limited to, naming, ascribing personality and gender. \cite{sung_housewives_2008} carried out a survey among 379 Roomba users to investigate how common these activities are.

We also found that people tend to anthropomorphize or interact socially (in a broader sense) with their vacuum cleaning robot.
However, our long-term observations suggest that this effect is marginal, and that the social impact of a functional robot on a household seems to be generally overestimated.
In contrast to the previous work, we found that there were just few instances of people anthropomorphizing the robot and interacting with it in a human-social way. More importantly, overall, these anthropomorphic interactions with the robot did not last but seemed to be somehow related to the novelty of the robot.

In terms of how people interacted with the robot, we categorized several interactions and actions with the robot as anthropomorphic or social, in a broader sense. Examples of what we categorized as somehow social and anthropomorphic (human-like) interactions with the robot include \textit{watching} the robot for fun, giving a \textit{demo} of the robot to guests / visitors, taking a \textit{photo/video} of the robot, \textit{personalizing} the robot, giving a \textit{name} to the robot, \textit{talking} to the robot using direct speech, \textit{playing} with the robot, and making \textit{adjustments} to the home due to the robot. Figure \ref{fig:roomba-activities}) shows how many of the 9 households engaged in what kind of activity with the robot at three different points of time after the robot was deployed in the household.

\paragraph*{Watching the robot for fun.} We categorize watching a robot for fun as an entertaining, social activity, reflecting a feeling of empathy for the robot. In \cite{sung_housewives_2008} survey, 73~\% of the participants reported to watch their Roomba running for fun. While in the beginning of our study, also a quite high proportion (6 of 9 households) reported to watch the robot for fun, after 2 and 6 months the number of households that engaged in this activity decreased by half.

\paragraph*{Giving a robot demo to visitors.} \cite{sung_domestic_2010} described that a vacuum cleaning robot can serve as a mediator and encourage social interaction among people. We categorize giving a demo of the robot to visitors / guest to the household as such an activity. During the first 2 weeks with the robot 8 of 9 households in our study showed their robot to others, and still after 2 and 6 months 5 of the households engaged in this activity. This number is more or less in accordance with the 57~\% of survey respondents who stated to give demonstrations to others \citep{sung_housewives_2008}.

\paragraph*{Taking photos / videos of the robot.} How far taking a photo or video of a robot can be seen as an anthropomorphic interaction is questionable. However, it reflects the fact that the robot is observed by the user who for some reason is motivated to record the event. The number of households that took a photo or video of the robot decreased over time from 5 households in the beginning to 1 household after 6 months.

@@

\paragraph*{Playing with the robot.} In our study, it were mostly children who engaged in playing with the robot, however, most of them stopped doing so after some time. The number of household playing with their Roomba decreased from 8 when asked after 2 weeks with the robot, to 4 households after 2 months, and 1 household after 6 months. In \cite{sung_housewives_2008} survey, 37~\% of the Roomba owners reported to play and experiment with their robot.

\paragraph*{Making adjustments to the home due to the robot.} To help Roomba clean, \cite{sung_housewives_2008} found in an online survey that about half of the 379 participants made physical modifications to their homes. This process is referred to as \textit{roombarization}, and it seems to be a unique aspect of adopting a domestic service robot. In our study, almost all the households made adjustments when they used the robot for the first time in front of us. However, during the first two weeks with the robot, only 3 households reported to have adjusted something due to the robot, and after 6 months it was only 1 household that still engaged in this activity. This decrease is, however, logical since after some time most people had done all the necessary adjustments to enable the robot show a good performance.

\paragraph*{Emerging anthropomorphic interactions.} Another, from our side more unexpected interaction that we would describe as somehow anthropomorphic, is what we call \textit{``collaborative cleaning''} with the robot. In the initial phase of the study, some people spontaneously helped the robot to clean by either picking up crumbs from the floor themselves or pushing it right in front of the robot. Another way of collaborating with the robot we observed was to help it go everywhere by spontaneously putting up the trash bin or the carpet in the bathroom (things that are not just lying around and need to be cleaned up). However, the activity of collaborative cleaning faded out soon, and we suspect it was encouraged by the novelty of the robot.

\subsubsection{Perception of human-like characteristics of the robot}

@@

    \subfloat[activities with the robot over time,\\ n = 9 households]
    {\label{fig:roomba-activities}
    \includegraphics[width=8.5cm]{roomba-activities.png}}
	\subfloat[people's perception of the robot over time,\\ n = 15 participants]    
	{\label{fig:roomba-perception}
    \includegraphics[width=8.5cm]{roomba-perception.pdf}}

    \caption{\textbf{Activities with and perception of the robot over time.}\\
	\textbf{(a)} This bar chart shows how many of the 9 households carried out different kinds of anthropomorphic / social activities with and related to the robot at 3 time points during 6 months. Activities include \texttt{watching} the robot for fun, giving a \texttt{demo} of the robot to guests / visitors, taking a \texttt{photo/video} of the robot, \texttt{personalizing} the robot, giving a \texttt{name} to the robot, \texttt{talking} to the robot, \texttt{playing} with the robot, and making \texttt{adjustments} to the home due to the robot. Participants were asked to report on these activities at 3 time points after the robot was deployed in their home: after 2 weeks, after 2 months, and after 6 months.\\
	\textbf{(b)} This line chart shows how people's perception of the robot evolves over time. At 5 time points people were asked to rate the robot according to the following aspects: the perceived \texttt{intelligence} of the robot, the perceived \texttt{usefulness}, the perceived \texttt{easy of use}, how much \texttt{fun} it was to use the robot, how far they found the robot had an \texttt{impact} on their household, and their {overall} impression of it.}

@@

In accordance with \cite{sung_housewives_2008} we also found relatively low rates of users treating and perceiveing their vacuum cleaning robot as a living (or human-like) entity. We interpret that it might be the pure functional purpose and non human-like form of the Roomba vacuum cleaning robot that prevented people from anthropomorphizing it. Still, we observed some initial anthropomorphic effects, such as collaborative cleaning with the robot, and talking to the robot.

@@

\hl{Julia: I think we should also include this study}

>>>>

We openly explored long-term interactions between human users and a domestic service robot in a 6-months lasting ethnographic study. We gave a commercially available vacuum cleaning robot (iRobot's Roomba) to nine households in the French speaking part of Switzerland. The fairly diverse sample of households consisted of 2 single homes, 1 couple, and 6 families with children in their home. In total, this summed up to 30 participants: 15 adults (26-71 years old, mean age 43.6 years), and 15 children (6 months to 18 years old). During the 6 months, each household was visited 5 times to conduct interviews and make field observations in the households. Moreover, during several weeks, households were asked to fill in a diary to report the usage of the robot and other cleaning activities.
\footnote{This study is published in \cite{fink_living_2013}.}
Hereafter, we refer to this study as the "Roomba study".

%The main goal of this study was to explore how households live with a vacuum cleaning robot over an extended period of time. We wanted to analyze how the usage of and interaction with the robot evolve over time, and to identify adoption phases as well as factors that impact on the long-term acceptance of service robots in homes.

From the Roomba study, we can report on several findings related to anthropomorphism and the anthropomorphic design of robots.
%The gathered data for these aspects include qualitative observations, semi-structured interviews, and also quantitative assessments of various aspects of human-robot interaction.
The Roomba study is one of very few long-term human-robot interaction studies in the domestic environment.

In the following, we present findings in respect of two aspects:
First, we qualitatively describe what kinds of anthropomorphic interactions between participants and the robot were observed and outline how these interactions evolved over time.
Second, we present people's perception of human-likeness and social capabilities of the robot and how this perception changed over time.\footnote{\hl{note that these two different kinds of data (behavior/interaction and perception) already show that we try to measure anthropomorphism from these two angles!}}

\subsubsection{Anthropomorphic interactions with the robot\\}

It has been described in several previous studies of the Roomba that users tend to engage in a variety of non-cleaning activities associated with the robot, including, but not limited to, naming, ascribing personality and gender \citep{forlizzi_service_2006,forlizzi_how_2007,krumm_my_2007,sung_domestic_2010}. \cite{sung_housewives_2008} carried out a survey among 379 Roomba users to investigate how common these activities are. The authors concluded that overall, respondents had relatively low rates of treating Roomba as a living entity.

In our long-term study of the Roomba in homes, we found that it was mostly during the initial phase with the robot that some (rather few) people tended to interact in an anthropomorphic or human-social way with their vacuum cleaning robot.
Overall, our data suggest that the anthropomorphic effect and the social impact of a functional robot on a household is rather marginal.
%There were just few instances of people anthropomorphizing the robot and interacting with it in a human-social way. More importantly, overall, these anthropomorphic interactions with the robot did not last but seemed to be somehow related to the novelty of the robot.

We categorized several interactions and actions with the robot as anthropomorphic or human-social (in a broader sense). 
%Examples of what we categorized as somehow social and anthropomorphic (human-like) interactions with the robot include \textit{watching} the robot for fun, giving a \textit{demo} of the robot to guests / visitors, taking a \textit{photo/video} of the robot, \textit{personalizing} the robot, giving a \textit{name} to the robot, \textit{talking} to the robot using direct speech, \textit{playing} with the robot, and making \textit{adjustments} to the home due to the robot. 
Figure \ref{fig:roomba-activities}) shows how many of the 9 households engaged in these activities during the 6 months when the study was conducted.

\paragraph*{Watching the robot for fun.} We categorize watching a robot for fun as an entertaining, emotional activity, reflecting a feeling of empathy for the robot. In \cite{sung_housewives_2008} survey, 73~\% of the participants reported to watch their Roomba running for fun. While in the beginning of our study, also a quite high proportion (6 of 9 households) reported to watch the robot for fun, after 2 and 6 months the number of households that engaged in this activity decreased by half.

\paragraph*{Giving a robot demo to visitors.} It is somehow natural that owners of innovative devices show the device to others and ask for their opinion about it \citep{rogers_diffusion_1995}. However, this activity suggests that the robot is presented as a new part of the household to visitors. \cite{sung_domestic_2010} described that a vacuum cleaning robot can serve as a mediator and encourage social interaction among people. We categorize giving a demo of the robot to visitors / guests as such an activity. During the first 2 weeks with the robot 8 of 9 households in our study showed their robot to others, and still after 2 and 6 months 5 of the households engaged in this activity. This number is more or less in accordance with the 57~\% of survey respondents who stated to give demonstrations to others \citep{sung_housewives_2008}.

%\paragraph*{Taking photos / videos of the robot.} How far taking a photo or video of a robot can be seen as an anthropomorphic interaction is questionable. However, it reflects the fact that the robot is observed by the user who for some reason is motivated to record the event. The number of households that took a photo or video of the robot decreased over time from 5 households in the beginning to 1 household after 6 months.

@@

\paragraph*{Playing with the robot.} In our study, it were mostly children who engaged in playing with the robot, however, most of them stopped doing so after some time. The number of household playing with their Roomba decreased from 8 when asked after 2 weeks with the robot, to 4 households after 2 months, and 1 household after 6 months. In \cite{sung_housewives_2008} survey, 37~\% of the Roomba owners reported to play and experiment with their robot.\\

%\paragraph*{Making adjustments to the home due to the robot.} To help Roomba clean, \cite{sung_housewives_2008} found in an online survey that about half of the 379 participants made physical modifications to their homes. This process is referred to as \textit{roombarization}, and it seems to be a unique aspect of adopting a domestic service robot. In our study, almost all the households made adjustments when they used the robot for the first time in front of us. However, during the first two weeks with the robot, only 3 households reported to have adjusted something due to the robot, and after 6 months it was only 1 household that still engaged in this activity. This decrease is, however, logical since after some time most people had done all the necessary adjustments to enable the robot show a good performance.

%\paragraph*{Emerging anthropomorphic interactions.} Another, from our side more unexpected interaction that we would describe as somehow anthropomorphic, is what we call \textit{``collaborative cleaning''} with the robot. In the initial phase of the study, some people spontaneously helped the robot to clean by either picking up crumbs from the floor themselves or pushing it right in front of the robot. Another way of collaborating with the robot we observed was to help it go everywhere by spontaneously putting up the trash bin or the carpet in the bathroom (things that are not just lying around and need to be cleaned up). However, the activity of collaborative cleaning faded out soon, and we suspect it was encouraged by the novelty of the robot.\\

Overall, Figure \ref{fig:roomba-activities} shows that as time goes by, less households engage in the before-mentioned non-cleaning activities with the vacuum cleaning robot: after 2 weeks, on average, 4.67 of 9 households engage in a human-like activity with the robot; after 2 months, the value decreases to 2.83, and after 6 months to 2.3. Our interpretation is that the nature of human-robot interaction changes over time. More concretely, we suspect  that anthropomorphic / human-social interactions with the robot become less over time, and that only with some people these interactions sustain.

@@

    \subfloat[activities with the robot over time,\\ n = 9 households]
    {\label{fig:roomba-activities}
    \includegraphics[width=8.5cm]{roomba-activities.pdf}}
	\subfloat[people's perception of the robot over time,\\ n = 15 participants]    
	{\label{fig:roomba-perception}
    \includegraphics[width=8.5cm]{roomba-perception.pdf}}

    \caption{\textbf{Activities with and perception of the robot over time.}\\
	\textbf{(a)} This bar chart shows how many of the 9 households carried out different kinds of anthropomorphic / social activities with and related to the robot at 3 time points during 6 months. Activities include \texttt{watching} the robot for fun, giving a \texttt{demo} of the robot to guests / visitors, \texttt{personalizing} the robot, giving a \texttt{name} to the robot, \texttt{talking} to the robot, \texttt{playing} with the robot. Participants were asked to report on these activities at 3 time points after the robot was deployed in their home: after 2 weeks, after 2 months, and after 6 months.\\
	\textbf{(b)} This line chart shows how people's perception of the robot evolves over time. At 5 time points people were asked to rate the robot according to the following aspects: the perceived \texttt{intelligence} of the robot, the perceived \texttt{usefulness}, the perceived \texttt{ease of use}, how much \texttt{fun} it was to use the robot, how far they found the robot had an \texttt{impact} on their household, and their {overall} impression of it.}

@@

\subsubsection{Perception of anthropomorphic characteristics of the robot\\}

We also studied people's perception of the robot, and how it evolved over time (see Figure \ref{fig:roomba-perception}). The main idea was to find out which of the perceived attributes could be related to how far the robot was integrated in the household as a cleaning tool.

Concerning anthropomorphism, we included \textit{perceived intelligence} and experienced \textit{fun} when using the robot as indicators for the perception of human-likeness in the robot.

\paragraph*{Perceived intelligence.} On a scale from 1-7, people were not sure about how intelligent they should expect the robot to be (mean value 3.9). The rating did not change much and not significantly over time (average ratings between 3.3-4.2) \citep{fink_living_2013}. Interestingly, at all five time points, male participants rated the robot as more intelligent (average 4.5) than females (average 3.4), however, this difference was not significant. When asked why they thought Roomba possesses intelligence, men more often referred to the 'robot side' of Roomba: its abilities to detect and avoid obstacles, stairs, and being able to recharge itself. Contrary, women referred to its rather limited vacuuming performance and found it was not as intelligent because \textit{``it doesn't see where the dirt is, it sometimes leaves it, it can't be smart!''}. This difference in the perception of intelligence of a robot suggests that gender,and probably other individual characteristics need to be taken into account.

\paragraph*{Experienced fun.} Participants rated the robot as 'somewhat fun' (values from 4.0-4.6). There was a qualitative gender difference (not significant), in the sense that the experienced fun remained quite stable for male participants, whereas it tended to decrease with time for female participants. Men described the robot more as a 'fun gadget' than females, who referred to it as \textit{``In the end, Roomba is a cleaning tool and it is not really fun to use it''}. Again for this aspect, gender seems to play a role in how the robot is perceived and experienced.

\subsubsection{Summary Roomba study\\}

In accordance with \cite{sung_housewives_2008} we also found relatively low rates of users treating and perceiving their vacuum cleaning robot as a living (or human-like) entity. We interpret that it might be the pure functional purpose and non human-like form of the Roomba that prevented people from anthropomorphizing it. Still, we observed some initial anthropomorphic effects, both reflected in interaction with and perception of the robot. However, most of the anthropomorphic effects did not last long. \hl{One fairly stable human-like interaction with the robot was \textit{talking to the robot}, despite the fact that the robot was not able to recognize and respond to verbal statements.\footnote{important! I think I've seen that finding somewhere else when comparing ASIMO to AIBO or so.}}

@@

\hl{Julia: I think we should also include this study ... if it is not getting too much}
